---
title: "Modeling Population Decline of Endangered Species"
author: "Peter Boyd"
date: "11/26/2018"
output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
subtitle: 'Link to Git repo: https://github.com/ST541-Fall2018/boydpe-project-populationdecline'
--- 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(MASS)
library(broom)
whale <- read_csv(here("data", "sperm_whale_pop.csv"))
monarch <- read_csv(here("data", "monarch.csv"))
addax <- read_csv(here("data", "addax.csv"))
grizzly_ukraine <- read_csv(here("data", "grizzly_ukraine.csv"))
ringed_seal <- read_csv(here("data", "ringed_seal_finland.csv"))
monarch$Population <- monarch$Population * 50000000
```

```{r, echo = FALSE}

pop.decline <- function(df, ntimes){
  df <- na.omit(df)
  fit.glm <- glm(df$Population ~ df$Year, data = df)
  fit.poi <- glm(df$Population ~ df$Year, data = df, family = "poisson")
  fit.nb <- glm.nb(df$Population ~ df$Year, data = df)
  #fit.log <- glm(Population ~ log(Year), data = df)
  fits <- list((fit.glm), (fit.poi), (fit.nb))
  #lm.fit <- lm(log(Population) ~ log(Year), data = df)
  
  #compare AIC
  mods <- c("Linear Model", "Poisson", "Negative Binomial")#, "Log Transformation")
  aics <- c(summary(fit.glm)$aic, summary(fit.poi)$aic, summary(fit.nb)$aic)
            #summary(fit.log)$aic)
  best_model <- fits[[order(aics)[1]]]
  aic_summary <- cat("AIC for Each Model", "\n",
               mods[1], ": ", aics[1], "\n", 
               mods[2], ": ", aics[2], "\n",
               mods[3], ": ", aics[3], "\n")
              # mods[4], ": ", aics[4], "\n")
  
  
  # do simulation for lowest aic
  n <- max(df$Year) - min(df$Year) +1
  y.hat <- best_model$fitted.values
  
  if(order(aics)[1] == 2){
    y.rep <- matrix(ncol = ntimes, data = c(rpois(n*ntimes, y.hat)))
  } else if(order(aics)[1] == 3){
    y.rep <- matrix(ncol = ntimes, data = c(rnegbin(n = n*ntimes, mu = y.hat, 
                  theta = summary(fit.nb)$theta)))
  # } else if(order(aics)[1] == 4){
  #   new_data <- data.frame("Year" = seq(max(df$Year)+1, max(df$Year) + n*ntimes, 1))
  #   preds <- predict(fit.log, newdata = new_data, type = 'response')
  #   preds <- preds + runif(n = n*ntimes, min = min(fit.glm$residuals), max =  
  #             max(fit.glm$residuals))
  #   y.rep <- matrix(ncol = ntimes, data = preds) 
  } else {
    new_data <- data.frame("Year" = seq(max(df$Year)+1, max(df$Year) + n*ntimes, 1))
    preds <- predict(fit.glm, newdata = new_data, type = 'response')
    preds <- preds + runif(n = n*ntimes, min = min(fit.glm$residuals), max =  
              max(fit.glm$residuals))
    y.rep <- matrix(ncol = ntimes, data = preds)
  }
    
  years <- c(min(df$Year):max(df$Year))
  y.rep.sort <- -apply(-y.rep, 2, sort)
  y.coefs <- t(apply(y.rep.sort, 2, function(y.col) lm(y.col~years)$coef))
  end.dates <- -y.coefs[,1] / y.coefs[,2]
  end.date <- cat("Predicted Extinction: ", mean(end.dates))
  
  #show plot with each model fit
  plot1 <- ggplot(data = df, aes(x = Year, y = Population)) +
  geom_point() +
  geom_smooth(method='glm',formula=y~x, method.args = list(family = "poisson"), 
                 aes(color = "Poisson")) +
  geom_smooth(method='glm',formula=y~x, aes(color = "Linear")) + 
  geom_smooth(method='glm.nb',formula=y~x, aes(color = "Neg.Bi")) +
  #geom_smooth(method='glm',formula=y~log(x), aes(color = "Log.Trans")) +
  scale_colour_manual(name="legend", values=c("orange", "green", "purple")) +
  theme(legend.position="bottom") + 
  #theme(legend.position = c(1,1)) +
  theme(legend.text = element_text(size = 5)) + 
  theme(legend.title = element_text(size = 8)) +
  guides(fill=guide_legend(title="Model")) + 
  scale_x_continuous(limits = c(round(min(df$Year)), round(max(df$Year)))) + 
  labs(x = "Year", y = "Population")
  
  #show residual plot for lowest aic model
  aug.best_model <- augment(best_model)
  resid.plot1 <- ggplot(aug.best_model, aes(x = .fitted, y = .resid)) +
    geom_point()
  
  print_plots <- function(x, y){
    {par(mfrow = c(1,2))
    print(x)
    print(y)}
  }
  
  #return stuff
  return(list(aic_summary, end.date, print_plots(plot1, resid.plot1)))
}
```



## Data and Implementation

- Data: any dataset containing columns titled "Year" and "Population"
- Function steps:
    1) Fits several different models
    2) Selects the best model by comparing AIC values
    3) Using fitted values from the best model, simulate many versions of new data
    4) Refit glm's, using original year values and simulated population
    5) Average fits, roughly estimate when population is zero
    5) Show graphically the fit of various models and the residual plot of best model
- `pop.decline(df = whale, ntimes = 100)`



## Example for Whale data set

```{r, echo=FALSE, fig.height=2.5, fig.width=2.4}
out <- pop.decline(df = whale, ntimes = 100)
out
#out$aic_summary
#print(out$plot1)
# {par(mfrow = c(1,2))
# print(pop.decline(df = whale, ntimes = 100)[[3]])
# print(pop.decline(df = whale, ntimes = 100)[[4]])}
```


## Example for Addax data set

```{r, echo=FALSE, message = FALSE, warning=FALSE, fig.height=2.5, fig.width=2.4}
pop.decline(df = addax, ntimes = 100)
```

## Ideas for Future Work

- Could add complexity to model by considering a carrying capacity variable in the function
- Attempted to iteratively predict new values until population is extinct

- Questions?

