---
title: "Week 7 Tasks"
author: "Peter Boyd"
date: "11/9/2018"
output: html_document
---

I'm not sure how well my whale data will work. I think I may have more success with this monarch data that I found. 

https://blog.nwf.org/2017/02/new-numbers-show-monarch-butterfly-populations-still-in-trouble/

http://www.columbia.edu/~cjd11/charles_dimaggio/DIRE/resources/R/simRreg.pdf

Towards bottom of this page, nice graph. Should include several lines that each show a different fit of the data, offering comparisons into the types fo regression. 

https://stats.idre.ucla.edu/r/dae/negative-binomial-regression/


```{r}
library(tidyverse)
library(here)
monarch <- read_csv(here("data", "monarch.csv"))
monarch
plot(monarch$Year, monarch$Acres, main = "Population Decline of Monarch Butterflies", xlab = "Year", ylab = "Hectacre Coverage")
```


Below, `pop.decline()` is function based on the poisson distribution. 

```{r}

# Need to make some changes. 
# Add if statements to separate cts and discrete data
# for discrete, we'll do classic glm, poi, nb, nls
# for cts, we'll do classic glm, exp, beta
# may be helpful to export:
#   plot with each fit 
#   aic values
#   residual plots
#   maybe make reco based on aic, subjective

pop.decline <- function(df, ntimes){
  fit.poi <- glm(df$Population ~ df$Year, data = df, family = "poisson")
  #fit.nb <- glm.nb(df$Population ~ df$Year, data = df)
  #find way to select based on min aic
  n <- length(df$Population)
  X <- cbind(rep(1, n), df$Year)
  b <- coef(fit.poi)
  y.hat <- fit.poi$fitted.values #exp(X %*% b)
  
  coefs <- matrix(ncol = 3, data = rep(0, ntimes*3))
  y.rep <- matrix(ncol = ntimes, data = rpois(n*ntimes, y.hat))
  for (i in 1: ntimes){
    coefs[i, 1] <- summary(lm(y.rep[,i] ~ df$Year))$coefficients[1]
    coefs[i, 2] <- summary(lm(y.rep[,i] ~ df$Year))$coefficients[2]
    coefs[i, 3] <- summary(lm(y.rep[,i] ~ df$Year))$coefficients[3]
  }

  
  new.fit <- colMeans(coefs)[1] + colMeans(coefs)[2]*seq(max(df$Year), max(df$Year) + 50 , 1) + colMeans(coefs)[3]*I(seq(max(df$Year), max(df$Year) + 50 , 1))^2
return(list(colMeans(coefs)))
#yr.extinct <- -colMeans(coefs)[1]/colMeans(coefs)[2]
#return(list(yr.extinct, colMeans(coefs)))
}

pop.decline(df = monarch, ntimes = 100)
```


Below is a nonlinear regression method. 


```{r}
monarch <- read_csv(here("data", "monarch.csv"))
monarch$Year <- monarch$Year - min(monarch$Year) + 1
lm.fit <- lm(log(Population) ~ log(Year), data = monarch)
nls.fit <- nls(Population ~ b0*Year^b1, 
    data = monarch, start = list(b0 = exp(coef(lm.fit)[1]), b1 = coef(lm.fit)[2]))


ggplot(data = monarch, aes(x = Year, y = Population)) +
  geom_line() +
  geom_smooth (method = "nls",
               formula = y ~ b0*x^b1, 
    data = monarch, start = c(b0 = exp(coef(lm.fit)[1]), b1 = coef(lm.fit)[2])) +
  #geom_smooth(method='glm',formula=y~x, family = "poisson") +
  labs(x = "Year", y = "Population")

plot(predict(nls.fit))
#looks exp, so use that as simulation aspect. 
# approximate rate para as 1/mean(pred nls fit)
x <- rexp(30, rate = 1/mean(predict(nls.fit)))
plot(sort(x, decreasing = T))

many.samples <- rexp(50*1000, rate = 1/mean(predict(nls.fit)))
msdf <- matrix(ncol = 1000, data = many.samples, byrow = F)

mins <- double(1000)
for (i in 1: 1000){
  mins[i] <- min(msdf[,i])  
}
sum(mins<0.01)/length(mins)
```

